Combinig geological and ecological proxies to model latitudinal temperature gradients in the early Eocene greenhouse

### 2022-12-02 manuscript/manuscript.Rmd
introduction

---
title: "An Early Eocene temperature gradient modelled with geological and ecological proxies"

@@

<!-- discarded titles:
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
-->

@@

# Introduction
The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current. Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern, potentially representing an analogue for extreme climate warming scenarios. Reconstructions of palaeoclimate commonly follow one of two approaches. Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions. Multiple local estimates are sometimes summarised to generate regional or global estimates, but this approach is prone to sampling bias [@Jones2022]. Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire glob.e This approach is, however, relying on proxy data to constrain the large number of possible model setups and ground truth the model output.

@@

Contrast between complex, process-driven Earth system models and very simplistic use of the proxy data as local temperature estimates or global averages. Here, we introduce a proxy-driven modelling approach of relatively low complexity, which places the available chemical and biological proxy data in a single climate model.
In order to properly compare model output to proxy data, latitudinal temperature patterns need to be reconstructed from noisy data (impossible to properly compare model gradients with scattered proxy points "by eye")
-->
# Materials & Methods

>>>>

---
title: "Combinig geological and ecological proxies to model latitudinal temperature gradients in the early Eocene greenhouse"

@@

<!-- discarded titles:
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
"An Early Eocene temperature gradient modelled with geological and ecological proxies"
-->

@@

# Introduction
The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021]. Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018]. Reconstructions of palaeoclimate commonly follow one of two approaches. Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g. @Liu2009]. To generate regional or global average estimates, the mean of multiple local estimates is taken [e.g. @Veizer2015]. This approach has the advantage of being transparent and easy to reproduce, modify or expand with new data, but is prone to sampling bias [@Jones2022]. Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire globe [@???]. However, this approach still relies on proxy data to constrain the large number of possible model setups and ground truth the model output [@???] It also requires substantial computing resources and technical expert knowledge, and is thus difficult to implement or modify for most palaeoclimate researchers.

Climate reconstructions of low - intermediate complexity provide an alternative that can be employed more readily by a broad range of users. Examples of this approach include proxy compilations that take into account the spatial nature of the data, e.g. @Hollis2019 investigated latitudinal temperature trends across the late Palaeocene - early Eocene with a spatially resolved proxy record. Expanding from this approach, some proxy-based reconstructions use statistical modelling to infer climatic patterns, and to bridge data gaps, e.g. using polynomials to reconstruct latitudinal gradients [@Bijl2009], or Gaussian process regression for a 2D-reconstruction of surface temperatures [@Inglis2020]. These latter approaches take advantage of the autoregressive nature

@@

Contrast between complex, process-driven Earth system models and very simplistic use of the proxy data as local temperature estimates or global averages. Here, we introduce a proxy-driven modelling approach of relatively low complexity, which places the available chemical and biological proxy data in a single climate model.
In order to properly compare model output to proxy data, latitudinal temperature patterns need to be reconstructed from noisy data (impossible to properly compare model gradients with scattered proxy points "by eye")

Reconstructing Earth's climate history is fundamental for understanding the evolution of the Earth system and its consequences for life in the past and future. The geological record, the only empirical source of information on the palaeoclimate of the deep past, holds evidence.

-->
# Materials & Methods
